Kindly Bent to Free Us

### 2019-07-04 kindly/motivation.tex
sudoku

The Sudoku solver maintains a 2D-array that represents the state of
the game and uses backtracking when there are several alternatives to proceed.
The use of backtracking suggests a persistent data structure for the 2D-array.
However, only changes that correspond to a choice point need to use
the persistence mechanism, others may be implemented as
cheap in-place mutations.

\cref{sig:hybarray} contains an API \lstinline/HybArray/ 
that guarantees correct use of mutable and immutable modifications to
the board by relying on affine types and borrows.
The signature differs slightly from the \lstinline/Array/
signature. For simplicity, the array elements are unrestricted, 
hence it is sufficent for the array to be affine.
The in-place mutation function is called \lstinline/set_mut/ with type
\lstinline/&!('a t) * int * 'a -> unit/ and the persistent
\lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
takes an \emph{immutable borrow} because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} is a very simple implementation of
\lstinline/HybArray/ that represents hybrid arrays
as regular arrays with copy-on-write for persistent
modifications. The function
\lstinline/create_from : int -> (int -> 'a) -> 'a Array.t /
initializes an array from a function.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

@@

Thanks to this API, modifications that correspond to backtracking points
use \lstinline/set/, which allows arbitrary uses of the previous version
of the array, while other modifications uses \lstinline/set_mut/, which
forbids uses of the previous version.
This can be seen in the Sudoku solver implementation in \cref{ex:sudoku}.
The board is represented as a matrix of cells (\cref{line:boardtype}).
We assume the existence of a \lstinline/Matrix/ module which uses the same API
as \lstinline/HybArray/ but with two indices.
Each cell is an integer
set that represent admissible solutions so far. The Sudoku solver will
iterate on the cells and try each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we use an immutable modification
to create a backtracking point (\cref{line:immutset}) and propagate
the changes with the \lstinline/propagate/ function. The \lstinline/propagate/
function does not need to preserve the previous
version of the board, and thus uses direct mutation.
The implementation of \lstinline/propagate/ is split into three parts
for lines, columns, and square. For concision, only \lstinline/propagate_lines/ is
presented here (\cref{line:propline}).
The typing ensures that all the mutations do not compromise
backtracking, since sharing is only allowed through
the immutable set functions, which is used for backtracking points.
In particular, the \lstinline/try_solution/ function can be used in a loop
precisely because the \lstinline/set/ function only requires an
\emph{immutable} borrow of \lstinline/new_g/.
This is allowed, since immutable borrows are unrestricted.

While we presented our API on simple copy-on-write arrays, this API
could easily be adapted to any persistent data-structure with
transient mutability such as RRB Vectors \cite{DBLP:journals/pacmpl/Puente17}
and persistent HAMT (used notably as Clojure's default associative maps
\cite{bagwell2001ideal,clojurehamt}) while providing a good performance profile
and a convenient programming style.

>>>>

The Sudoku solver maintains an array to represent the state of
the game and uses backtracking when there are several choices to proceed.
The use of backtracking suggests a persistent data structure for the array.
However, only changes that correspond to a choice point need to use
the persistence mechanism, others may be implemented as
cheap in-place mutations.

\cref{sig:hybarray} contains an API \lstinline/HybArray/ 
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the \lstinline/Array/
signature. For simplicity, the array elements are unrestricted, 
hence it is sufficent for the array to be affine.
The in-place mutation function \lstinline/set_mut/ with type
\lstinline/&!('a t) * int * 'a -> unit/ works on a mutable borrow  whereas the persistent
\lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
takes an immutable borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
\lstinline/HybArray/ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
\lstinline/create_from : int -> (int -> 'a) -> 'a Array.t /
initializes an array from a function.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

@@

Our implementation of a Sudoku solver (\cref{ex:sudoku}) performs
modifications that correspond to choice points using \lstinline/set/,
which makes it trivial to come back to the previous version
of the array, while other modifications use \lstinline/set_mut/, which
forbids uses of the previous version.

The board is represented as a 2D-matrix (\cref{line:boardtype}), where
the \lstinline/Matrix/ type uses the same API as \lstinline/HybArray/
but with two indices. 
Each cell contains an integer set that represent admissible solutions so far.
The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we create a choicepoint,
change the cell with an immutable modification (\cref{line:immutset}), and propagate
the changes with the \lstinline/propagate/ function.
The \lstinline/propagate/ function uses direct mutation through a
mutable borrow of the matrix as it need not preserve the previous
version of the board.  
The implementation of \lstinline/propagate/ is split into three parts
for lines, columns, and square, which are all very similar to function
\lstinline/propagate_lines/ (\cref{line:propline}).

The typing ensures that the mutations do not compromise the state at
the choice point, because they operate on a new state \lstinline/new_g/ created for one
particular branch of the choice.
In particular, the \lstinline/try_solution/ function can be used in a loop
precisely because the \lstinline/set/ function only requires an
unrestricted immutable borrow of \lstinline/g/, which can be used as
many times as needed.
We could even free \lstinline/new_g/ in \cref{line:free:new_g}.

While presented for copy-on-write arrays, the API
can easily be adapted to other persistent data structures with
transient mutability such as RRB Vectors \cite{DBLP:journals/pacmpl/Puente17}
or persistent HAMT \cite{bagwell2001ideal,clojurehamt} to provide  a
convenient programming style without compromising performance.
